Recurrent Deep Learning Models and its applications

### 2025-03-28 paper2/main.tex
paper update

% ######################## table of content ###############################
\tableofcontents % Generate a table of contents                        
% ######################## table of content ###############################

@@

This unfolds the RNN to construct a traditional Feedforward Neural Network where we can apply backpropagation through time. Below is the conceptual idea of BPTT in RNN.
\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{../Pic/pic1.png} % Replace "example-image" with your image file
    \caption{Unfolded RNN}
\end{figure}

@@

\subsubsection{Hidden Markov Model}
Before reviewing Hidden Markov Model (HMM), it is essential to understand what Markov models is, or Markov chains. Markov chains are fundamental models in probability theory and statistic, and it is a stochastic process describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. This property known as the Markov property \parencite{roberts2004general, rabiner1986anintroduction}. In rigorous terms. Let the state space be defined as:
\begin{equation}
    S = \{ s_1, s_2, \dots, s_N \}.
\end{equation}
\begin{equation}
    P(s_{t+1} \mid s_t, s_{t-1}, s_{t-2}, \cdots, s_{1}) = P(s_{t+1} \mid s_t)
\end{equation}
The state transition probability is denoted by:
\begin{equation}
    P_{ij} = P(s_{t+1} = s_j \mid s_t = s_i).
\end{equation}

@@

In this report, ACL-IMDB \parencite{maas-EtAl:2011:ACL-HLT2011} was experiment. This is a dataset for binary sentiment classification containing substantially more data than previous benchmark datasets. This dataset provide a set of 25,000 highly polar movie reviews for training, and 25,000 for testing. I preporcess the dataset by text cleaning and tokenization.

@@

\newpage
\section{Analysis}
\subsection{Learning Behavior}
All models show the effective learning during the initial epochs, roughtly 55\%, as both training and validation accuracy improve, and loss decrease. But the common problem is overfitting occure in 3 models. The possible reasons is dataset. Since every sentence in dataset was truncated. I just use the first 100 token to training the model. So the existing tokens might not be fully represent the sentence. The second reason is the embedding layers, I used the built-in function in Pytorch to create the embedding layer by randomise the matrix. 
\\[1ex]
Among the 3 models, LSTM and GRU have the better performance compare with vanilla RNN. The reason of GRU and LSTM perform better than RNN is the vanilla RNN struggles with long term dependencies because of the vanishing gradient problem when processing long sequences. In this experience is 100 time step. And this limits its ability to learn the meaningful patterns from data. On the contrast, GRU and LSTM use gate mechanisms to retain the relevant information and discard irrelevant information, allowing them to capture the long term dependencies more effectively.
\\[1ex]
And for the overfitting problem, we can see that overfitting problem occur in three models. For the vanilla RNN, since this is the basic and simple architecture to handle sequential data, it lacks mechanisms to retain or forget information selectively. As a result, it relying on the memorizing the training data rather than learning the meaningful patterns. This causes it to overfit. And for the GRU and LSTM, although gating mechanisms provide a more sophisticated architecture to learning from data, they aggressively optimizes the training loss, which can result in overfitting when the model continues to learn unnecessary information from the dataset. 
\\[1ex] 
The above result and analyse shown that GRU and LSTM perform better than vanilla RNN, which show that the graident vanishing and graident exploding problem were improved by the gating mechanism. And there is so many way will affect the experiment result for example, the embedding layer, tokenization.
\newpage

@@

% ######################## Reference ######################################
\printbibliography
% ######################## Reference ######################################

>>>>

% ######################## table of content ###############################
\tableofcontents % Generate a table of contents       
\newpage                 
% ######################## table of content ###############################

@@

This unfolds the RNN to construct a traditional Feedforward Neural Network where we can apply backpropagation through time. Below is the conceptual idea of BPTT in RNN.
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{../Pic/pic1.png} % Replace "example-image" with your image file
    \caption{Unfolded RNN}
\end{figure}

@@

\subsubsection{RNN encoder decoder model}
RNN Encoder-Decoderis a RNN based architecture which commonly used for handle sequence to sequence problem, for example machine translation, summarization, and question answering. The designed of this architecture is good for handle the input sequence and the output sequence is not the same length, leveraging the sequential nature of data. 
\\[1ex]
In this architecture, there are two main components, the first component is rnn encoder, the usage of the encoder is to encode the input sequence into a fixed length vector. And the decoder is to decode the vector into the target sequence. 
\\[1ex]
\textbf{Encoder}
\\[1ex]
The encoder is a RNN based block, which processes the input sequence step by step and calculate the hidden state representation for each input token. Let the input sequence be $X = [x_1, x_2, \cdots, x_T]$ where $T$ is the length of the input sequence, and $x_t$ is the input at time step $t$. We use the below formula to computes the hidden states $h_t$.
\\[1ex]
For the Vanilla RNN encoder
\begin{equation}
    h_t = f_{enc}(x_t, h_{t-1})
\end{equation}

where the $f_{enc}$ is the recurrent transition function, (e.g. LSTM, GRU or Vanilla RNN cell), and $h_t$ is the hidden state at time step $t$. And then the final hidden state of the encoder, $h_T$ will act as the context vector.
\begin{equation}
    c = h_T
\end{equation}
\textbf{Decoder}
\\[1ex]
The decoder is the second component of RNN encoder decoder, the usage of the decoder is to generates the output sequence step by step, conditioned on the context vector $c$ and its previous hidden state. 
\begin{equation}
    s_t = f_{dec}(y_{t-1}, s_{t-1}, c)
\end{equation}
where $f_{dec}$ is the recurrent function of the decoder, and it is just like the encoder part, $y_{t-1}$ is the embedding of the previous output token, $s_{t-1}$ is the decoder's hidden state. And $c$ is the context vector from the encdoer, which remains constant throughout decoding. 
\\[1ex]
After the model computed the $s_t$, the model will compute the output probability distribution.
\begin{equation}
    P(y_t|y_1, c) = softmax(W_os_t+b_o)
\end{equation}
And the predicted token $y_t = arg maxP(y_t)$, the process will repeat until the decoder predicts the end of sequence token $(<EOS>)$.

\subsubsection{Hidden Markov Model}
Before reviewing Hidden Markov Model (HMM), it is essential to understand what Markov models is, or Markov chains. Markov chains are fundamental models in probability theory and statistic, and it is a stochastic process, which describe a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. This property known as the Markov property \parencite{roberts2004general, rabiner1986anintroduction}. In rigorous terms. Let the state space be defined as:
\begin{equation}
    S = \{ s_1, s_2, \dots, s_N \}.
\end{equation}
\begin{equation}
    P(s_{t+1} \mid s_t, s_{t-1}, s_{t-2}, \cdots, s_{1}) = P(s_{t+1} \mid s_t)
\end{equation}
The state transition probability is denoted by:
\begin{equation}
    P_{ij} = P(s_{t+1} = s_j \mid s_t = s_i).
\end{equation}

@@

In this report, ACL-IMDB \parencite{maas-EtAl:2011:ACL-HLT2011} was use and for this dataset is a binary sentiment classification containing the substantially more data than the last version benchmark datasets. This dataset provide a set of 25,000 highly polar movie reviews for training, and 25,000 for validation. I preporcess the dataset by text cleaning and tokenization.

@@

\newpage
\section{Analysis}
\subsection{Learning Behavior}
All models show the effective learning during the initial epochs, roughtly 55\%, as both training and validation accuracy improve, and loss decrease. But the common problem is overfitting occure in 3 models. The possible reasons is dataset. Since every sentence in dataset was truncated. I just use the first 100 token to training the model. So the existing tokens might not be fully represent the sentence. The second reason is the embedding layers, I used the built-in function in Pytorch to create the embedding layer by randomise the matrix. 
\\[1ex]
Among the 3 models, LSTM and GRU have the better performance compare with vanilla RNN. The reason of GRU and LSTM perform better than RNN is the vanilla RNN struggles with long term dependencies because of the vanishing gradient problem when processing long sequences. In this experience is 100 time step. And this limits its ability to learn the meaningful patterns from data. On the contrast, GRU and LSTM use gate mechanisms to retain the relevant information and discard irrelevant information, allowing them to capture the long term dependencies more effectively.
\\[1ex]
And for the overfitting problem, we can see that overfitting problem occur in three models. For the vanilla RNN, since this is the basic and simple architecture to handle sequential data, it lacks mechanisms to retain or forget information selectively. As a result, it relying on the memorizing the training data rather than learning the meaningful patterns. This causes it to overfit. And for the GRU and LSTM, although gating mechanisms provide a more sophisticated architecture to learning from data, they aggressively optimizes the training loss, which can result in overfitting when the model continues to learn unnecessary information from the dataset. 
\\[1ex] 
The above result and analyse shown that GRU and LSTM perform better than vanilla RNN, which show that the graident vanishing and graident exploding problem were improved by the gating mechanism. And there is so many way will affect the experiment result for example, the embedding layer, tokenization.

\newpage

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/gradient.png}
\end{figure}
\subsubsection{Vanilla RNN (Red line)}
The gradients in the Vanilla RNN is unstable, as shown by the oscillations and spikes in the early timesteps. Over time, the gradients decay rapidly and go to 0, indicating that the presence of the vanishing gradient problem. This problem occurs because the Vanilla RNNs do not have the gating mechanisms in which this architecture is not perform well in long time dependencies. In other words, as the sequence length increases, repeated matrix multiplications lead to the gradients to shrink exponentially, making it difficult for the model to update the weighting matrix effectively during the BPTT. 
\\[1ex]
\subsubsection{LSTM (Blue line)}
The gradients of the LSTM are much more stable compared to the simple vanilla RNN. The gradient magnitudes remain consistent across timesteps and do not decay to zero. This stability is achieved through the use of gating machanisms that regulate the flow of information. These gates allow the LSTM to keep the relevant information form the input data while remove some irrelevant information from the data. As a result, LSTM is highly effective at learning long term dependencies in sequences.
\\[1ex]
\subsubsection{GRU (Green line)}
The gradients in the GRU are also stable, similar to the LSTM, but the performance is more smoother. GRU achieves this stability using its gate mechanism, which control the degree of information through the network. Compared to the LSTM, the GRU has a simpler architecture, which makes it more computationally efficient while still addressing the vanishing gradient problem.

\newpage

@@

% ######################## Discussion and Conclusion ######################################
\section{Discussion and Conclusion}

\subsection{Discussion}
During the experimental phase, several insights and challenges emerged, which are discussed below:

\begin{itemize}
    \item \textbf{Unexpected Results:}  
    It was observed that the loss of a two-layer LSTM model was higher compared to that of a one-layer LSTM. This was unexpected because additional layers are generally presumed to capture more complex patterns and improve performance. One possible explanation for this anomaly is the inherent randomness during model initialization and training. Since the performance difference can vary from run to run, it is possible that in some exceptions the two-layer model does not outperform the one-layer configuration.

    \item \textbf{Limitations:}  
    Due to limited computing resources, the experiments were constrained to training the models from scratch with relatively simpler architectures. This restricted the exploration of more complex or state-of-the-art models, such as Transformer-based architectures, which might offer further performance improvements but require substantially more computational power.
\end{itemize}

\subsection{Conclusion}
The primary objective of this report was to employ various RNN-based models and compare their performance on a sentiment classification task. The main findings of the project are:

\begin{itemize}
    \item \textbf{Comparative Analysis:}  
    Different RNN-based models, including vanilla RNN, LSTM, and GRU, were implemented and evaluated. The experiments show that LSTM and GRU models effectively address the long-term dependency problem inherent in sequential data. Their internal gating mechanisms allow them to control and preserve information over longer sequences.

    \item \textbf{Efficacy of Gated Architectures:}  
    The results indicate that the gated architectures (LSTM and GRU) significantly outperform the vanilla RNN in terms of handling vanishing and exploding gradient issues, thereby offering more stable and reliable performance.
\end{itemize}

In conclusion, while some unexpected results were observed—such as the higher loss in the two-layer LSTM model—the overall findings strongly advocate for the use of gated RNN architectures in tasks requiring modeling of long-term dependencies. Despite limitations due to computational resources, this report provides a valuable comparative study and lays the groundwork for future research on more complex models like Transformers in natural language processing tasks.

In conclusion, while the project revealed certain limitations, it nevertheless contributes valuable insights into the training and evaluation of RNN-based models for sentiment classification, thereby setting the stage for further research in the domain of efficient and scalable NLP methodologies.
% ######################## Discussion and Conclusion ######################################

% ######################## Reference ######################################
\newpage
\printbibliography
% ######################## Reference ######################################
